Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-04-22 analysis/paper/paper.Rmd
Minor adjustments to text

Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway that have been based on more informally driven methods, and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues.

@@

```{r wrangling}
sites <- assemblages %>% 
      # Find mean age for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch))) %>% 
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100)
```

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref@(fig:dist)A]. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects with assemblages holding more than 100 artefacts. The institution responsible for all of these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the Museum of Cultural History. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [@clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref@(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas. Furthermore, these coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.

The data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. This was combined with the blade category here. Furthermore, the artefact data has been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material accessibility and technological choices [@eigeland2015, 45-53]. Furthermore, a general sentiment is that an increase in the use of non-flint materials signifies an increased adaption to local environments, and possibly a reduction in mobility. However, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, the retained debitage and tool categories are expected to still reflect the differentiated use associated with different raw-materials. An important benefit of combining all non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised [cf. @frivoll2017].

@@

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016; @solheim2020], or in-depth analyses of handful of sites. Some studies have attempted more comprehensive evaluations of entire lithic assemblages associated with larger groups of sites, mainly to get at potential site types and associated mobility patterns among Early Mesolithic sites [@breivik2016; @viken2018a; @breivik2020]. However, only a single study has employed a multivariate quantitative framework in the treatment of the assemblages [cf. @naeroy2018, 905]. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to qualitatively and subjectively driven methods.

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is partially informed by the so-called Frison Effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes, and that the shapes studied by archaeologists are the shape they took upon discard. Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose, is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially entirely misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take into account, and renders it difficult to align any identified patterns with specific behavioural dimensions [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979]. Although precisely how degrees of retouch is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, as this is commonly seen as a means of extending the use-life or realising the potential utility of a tool by repeated rejuvination and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005].

The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies [e.g. @parry1987; @clark2017], as well as raw-material quality and availability [e.g. @andrefsky1994; @smith2015]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this framework assumed to lead to better control of raw-material availability, and to allow for the accumulation of material resources, which should in turn lead to an expedient technological organisation and reduce the necessity for extensive retouch. The material expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. Additional empirical expectations following from this is a higher number of cores and unretouched flakes and blades [@clark2017].

While the concept of curation and its empirical and behavioural implications have been widely discussed [e.g. @bamforth1986, @shott1996] , following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested. Through a series of studies, have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent and comparable results across a wide range of chronological and cultural context which are taken to reflect . This is the main measure employed here. Some additional variables suggested by @Bicho2020 to be related to mobility patterns are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures [cf. @baxter1994, 100].

@@

In sum, the results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. This is evident in the present data as well, highlighted by the influence of the later Mesolithic axe production sites, which it would be reasonable to assume corresponds to a more pervasive cultural change than a purely typo-chronological development of artefact shape. Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).

@@

```{r pcadata}
# Tried to do this to avoid having to type the variable names into both mutate()
# and select(). The general syntax select(-(vector_of_variable_names)) worked,
# but not: mutate(new_varaible = sum(vector_of_variable_names) / artefact_count)
primary_flint <- c("core", "core_fragment", "blade", "microblade", "chip",
                     "fragment", "flake")

primary_lithics <- c(primary_flint, "flake_other", "fragment_other", 
                     "chip_other", "blade_other",  "microblade_other", 
                     "core_other", "core_fragment_other")

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-primary_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count) %>% 
  column_to_rownames("site")

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for settlement strategies as operationalised by @clark2017 and @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data. hese last dimensions are capturing the most variation in the dataset presented here. Thus, if frequency of secondarily worked artefacts is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.

@@

The results of the CA align more or less with results of previous research that has considered the chronological . This would indicate that in general, meaningful patterning is associated with the employed artefact groups.

>>>>

However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway that have been based on more informally driven methods, and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

@@

```{r wrangling}
sites <- assemblages %>% 
      # Find mean age for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100)
```

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref\@(fig:dist)A]. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. @clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref\@(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas. Furthermore, these coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

The data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. This was combined with the blade category here. Furthermore, the artefact data has been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material quality and accessibility as well as technological choices [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of adzes and other macro tools. Given this differentiated use, the retained debitage and tool categories are expected to still reflect the differentiated use associated with different raw-materials. An important benefit of combining all non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised [cf. @frivoll2017].

The date ranges provided for the sites are taken from the original excavation reports, and are the result of a consideration of typological indicators, dating with reference to shoreline displacement, and radiocarbon dates where these are available. Employing these date-ranges was deemed sufficient, as the main purpose is to get at general tendencies in the material, and not fine-grained chronological developments.

@@

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g. @damlien2016; @solheim2020]. Some studies have involved in-depth analyses of handful of sites, either in terms of intrasite spatial organisation [@naeroy2018], or in terms of technological organisation. There are also some comprehensive evaluations of entire lithic assemblages associated with a larger group of sites to get at potential site types and associated mobility patterns [@breivik2016; @viken2018c; @breivik2020]. However, only a single study concerned with eight Middle-Mesolithic assemblages has employed a multivariate quantitative framework to help structure such a treatment [@naeroy2018, 905]. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively driven methods.

@@

The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is also partially informed by the so-called Frison effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take properly into account, and renders it difficult to align any identified patterns with specific behavioural dimensions [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979]. Although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, this is commonly seen as a means of extending the use-life, or realising the potential utility of a tool by repeated rejuvination and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies, as well as raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].

While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested. Through a series of studies @barton1999 and collegues [e.g. @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context. This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this framework assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. From this is a higher number of cores and unretouched flakes and blades [@clark2017].

This is the main measure employed here. Some additional variables suggested by @Bicho2020 to be related to mobility patterns are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures [cf. @baxter1994, 100].

@@

@@

```{r pcadata}
# Tried to do this to avoid having to type the variable names into both mutate()
# and select(). The general syntax select(-(vector_of_variable_names)) worked,
# but not: mutate(new_varaible = sum(vector_of_variable_names) / artefact_count)
blanks_flint <- c("blade", "microblade", "fragment", "flake")

blanks_lithics <- c(blanks_flint, "flake_other", "fragment_other", 
                     "blade_other",  "microblade_other")

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         blanks_freq = sum( blade, microblade, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
         -core_fragment, -chip) %>% 
  column_to_rownames("site")

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data. hese last dimensions are capturing the most variation in the dataset presented here. Thus, if frequency of secondarily worked artefacts is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.

@@

The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. This is evident in the present data as well, highlighted by the influence of the later Mesolithic axe production sites, which it would be reasonable to assume corresponds to a more pervasive cultural change than a purely typo-chronological development of artefact morphology. Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).

The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking. While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pair is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous. In attempt to determine if this might be the case, it was decided to introduce site locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020. Of the variables considered, this study found that the most consistently important environmental covariates for the location of sites in the region was related to the exposure of the sites, and failed to identify any diachronic variation in this pattern. As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similarly temporary , one could envisage that variablity in site functionality might be reflected in degree of exposure. However, introducing these variables does not clarify the picture, as both variables show negligible degrees of correlation with any of the variables pertaining to the assemblage data.
